---
title: "Vis_with_ggplot2"
author: "Darwin Keung"
date: "9/27/2018"
output: github_document
---
# Visualization with ggplot2
## Part 1

Good visualization is a critical step in data analysis.
This is the first module in the Visualization and EDA topic.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
## 
## Attaching package: 'ggridges'
## The following object is masked from 'package:ggplot2':
## 
##     scale_discrete_manual
```

```{r rnoaa setup}
library(rnoaa)

weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())
weather_df
```

### _Basic scatterplot_

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax))
```

Well, my “scatterplot” is blank. That’s because I’ve defined the data and the aesthetic mappings, but haven’t added any geoms:  ggplot knows what data I want to plot and how I want to map variables, but not what I want to show. Below I add a geom to define my first scatterplot.
```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point()
## Warning message: Removed 15 rows containing missing values (geom_point). 
```

The code below could be used instead to produce the same figure. Using this style can be helpful if you want to do some pre-processing before making your plot but don’t want to save the intermediate data.
```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```

Notice that we try to use good styling practices here as well – new plot elements are added on new lines, code that’s part of the same sequence is indented, we’re making use of whitespace, etc.

You can also save the output of ggplot() to an object and modify / print it later. This is often helpful, although it’s not my default approach to making plots.

```{r}
plot_weather = 
  weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) 

plot_weather + geom_point()
```

### _Advanced scatterplot_

The basic scatterplot gave some useful information – the variables are related roughly as we’d expect, and there aren’t any obvious outliers to investigate before moving on. We do, however, have other variables to learn about using additional aesthetic mappings.

Let’s start with `name`, which I can incorporate using the `color` aesthetic:
```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name))
## Warning: Removed 15 rows containing missing values (geom_point).
```

Nice! We get colors and have a handly legend. Maybe next I’ll add a smooth curve and make the data points a bit transparent.

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) +
  geom_smooth(se = FALSE)
## `geom_smooth()` using method = 'gam' and formula 'y ~ s(x, bs = "cs")'
## Warning: Removed 15 rows containing non-finite values (stat_smooth).
## Warning: Removed 15 rows containing missing values (geom_point).
```

Neat! The curve gives a sense of the relationship between variables, and the transparency shows where data are overlapping. I can’t help but notice, though, that the smooth curve is for all the data but the colors are only for the scatterplot. Turns out that this is due to where I defined the mappings. The X and Y mappings apply to the whole graphic, but color is currently geom-specific. Sometimes you want or need to do this, but for now I don’t like it. If I’m honest, I’m also having a hard time seeing everything on one plot, so I’m going to add facet based on name as well.

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) 
## `geom_smooth()` using method = 'loess' and formula 'y ~ x'
## Warning: Removed 15 rows containing non-finite values (stat_smooth).
## Warning: Removed 15 rows containing missing values (geom_point).
```

If I’m honest, I’m also having a hard time seeing everything on one plot, so I’m going to add facet based on name as well.

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
## `geom_smooth()` using method = 'loess' and formula 'y ~ x'
## Warning: Removed 15 rows containing non-finite values (stat_smooth).
## Warning: Removed 15 rows containing missing values (geom_point).
## ~ name: means that there is a new grid or "facet" based on the variables
```  

More interesting plot

```{r}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)  
```

geom_line vs geom_point line connects the dots

## Tips and tricks
There are lots of ways you can mix and match elements, depending on your goals. I don’t like the following plot as much (it doesn’t show the data and omits precipitation), but it illustrates that you have lots of freedom in determining which geoms to include and how to compare your groups. If nothing else, you should be starting to get a sense for why you create way more plots for yourself than for others.
```{r}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_smooth(se = FALSE) 
## `geom_smooth()` using method = 'loess' and formula 'y ~ x'
## Warning: Removed 3 rows containing non-finite values (stat_smooth).
```

When you’re making a scatterplot with lots of data, there’s a limit to how much you can avoid overplotting using alpha levels and transparency. In these cases `geom_hex()`, `geom_bin2d()`, or geom_density2d() can be handy:

```{r}
ggplot(weather_df, aes(x = tmax, y = tmin)) + 
  geom_hex()
## Warning: Removed 15 rows containing non-finite values (stat_binhex).
```

There are lots of aesthetics, and these depend to some extent on the geom – color worked for both `geom_point()` and  `geom_smooth()`, but `shape` only applies to points. The help page for each geom includes a list of understood aesthetics.

*Learning Assessment:* In the preceding, we set the alpha aesthetic “by hand” instead of mapping it to a variable. This is possible for other aesthetics too. To check your understanding of this point, try to explain why the two lines below don’t produce the same result:

```{r}
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin), color = "blue")
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin, color = "blue"))
```

In the first attempt, we’re defining the color of the points by hand; In the second attempt, we’re implicitly creating a color variable that has the value blue everywhere; ggplot is then assigning colors according to this variable using the default color scheme.

## Univariate plots

Scatterplots are great, but sometimes you need to carefully understand the distribution of single variables – we’ll tackle that now. This is primarily an issue of learning some new geoms and, where necessary, some new aesthetics.

First up is the histogram.

```{r}
ggplot(weather_df, aes(x = tmax)) + 
  geom_histogram()
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
## Warning: Removed 3 rows containing non-finite values (stat_bin).
```

You can play around with things like the bin width and set the fill color using an aesthetic mapping.

```{r}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2)
## Warning: Removed 3 rows containing non-finite values (stat_bin).
```

The `position = "dodge"` places the bars for each group side-by-side, but this gets sort of hard to understand. I often prefer density plots in place of histograms.

```{r}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_density(alpha = .4, adjust = .5, color = "blue")
## Warning: Removed 3 rows containing non-finite values (stat_density).
```

mac in RStudio: alt+cmd+i. It's under the "code" dropdown menu box if you forget it
